---
title: "HW_1"
format: html
editor: visual
---

Used libraries

```{r}
library(scales)
library(ggplot2)
library(dplyr)
library(tidyr)
library(reshape2)
library(data.table)
library(DESeq2)
library(apeglm)
library(tibble)
library(ggfortify)
library(pheatmap)
library(RColorBrewer)
library(org.Hs.eg.db)
library(clusterProfiler)
library(msigdbr)
library(biomaRt)
library(enrichplot)
library(ggridges)
```
Loading Data


```{r}
counts_raw <- read.delim("C:/Users/pauli/Downloads/count_table.txt",
skip = 1)

counts_raw$Geneid <- gsub("\\..*$", "", counts_raw$Geneid)
dim(counts_raw)

colnames(counts_raw) <- c("Geneid",	"Chr",	"Start",	"End",	"Strand",
  "Length",	"SRR11647685",	"SRR11647687",	"SRR11647689",
  "SRR11647695",	"SRR11647697",	"SRR11647699")

colnames(counts_raw)
head(counts_raw)

```

\`

```{r}
meta <- read.table(
  "C:/Users/pauli/Downloads/sampleInfo.txt",
  header = TRUE,
  sep = " "
)
print(meta)
```

```{r}
gene <- read.table(
  "C:/Users/pauli/Downloads/gene_id_type.tsv",
  header = TRUE
)
colnames(gene) <- c("Geneid",	"Gene_Type")
colnames(gene)
dplyr::count(gene, Gene_Type)


gene$Geneid <- substr(gene$Geneid, 1 ,15)
head(gene)


gene_filter <- gene %>% filter(Gene_Type == 'protein_coding')

head(gene_filter)
dim(gene_filter)

```

```{r}
counts_filter <- counts_raw %>% dplyr::select(-2, -3, -4, -5, -6)
head(counts_filter)

```


```{r}
counts_gene <- left_join(gene_filter, counts_filter, by = "Geneid")
counts_gene <- counts_gene  %>% dplyr::select(-2)

sum(is.na(counts_gene))
dim(counts_gene)
```

```{r}

num_gene <- colSums(counts_gene[-1])
meta$Counts <- num_gene
```

Library Size Exploration

```{r}

ggplot(meta, aes(x=Sample, y=Counts, fill=Type)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = label_number(scale = 1e-6)) +
  labs(title = "Lybrary Sizes per Sample",
       subtitle = "Total Mapped Reads (raw counts, protein-coding genes",
       x = "Sample",
       y = "Counts (in Millions)")

```


Counting Distribution

```{r}
meta$log2n <- log2(meta$Counts+1)
summary(count_log)

meta$median <- log2 (c(202, 261, 250, 263, 197, 227)+1)
meta$Q1 <- log2(c(8, 14, 12, 10, 6, 6)+1)
meta$Q3 <- log2(c(738, 985, 908, 991, 799, 920)+1)

ggplot(meta, aes(x=Sample, y=log2n, fill=Type)) +
  geom_bar(stat = "identity") +
  labs(title = "Distribution of log2(counts + 1) per Sample",
       x = "Sample",
       y = "log2(counts + 1)")+
  

  geom_pointrange(aes(y = median,
                      ymin = Q1,
                      ymax = Q3),
                  show.legend = FALSE) +
  labs(       subtitle = 'Median and interquartile range',
       )
```

Subset to Normal vs Tumor
```{r}
samples_of_interest <- meta %>%
  filter(Type %in% c("Normal", "Tumor"))

print(samples_of_interest[, c("Sample", "Type", "Counts")])

counts_subset <- counts_gene[, samples_of_interest$Sample]
cat("\nDimensions of subset count matrix:", dim(counts_subset), "\n")
```

Building the DESeqDataSet

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = counts_subset,
  colData   = samples_of_interest,
  design    = ~ Type
)

dds$Type <- relevel(dds$Type, ref = "Tumor")
```

```{r}

keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]

cat("Genes remaining after low-count filter:", nrow(dds), "\n")

```
Running DESeq2

```{r}
dds <- DESeq(dds)


res <- results(dds,
               contrast = c("Type", "Normal", "Tumor"),
               alpha = 0.05)  

summary(res)

```

Applying Log2FC Shrinkage

```{r}
res_shrunk <- lfcShrink(dds,
                        coef = "Type_Normal_vs_Tumor",
                        type = "apeglm")

summary(res_shrunk)

```

Converting Results to a Data Frame

```{r}

res_df <- res_shrunk %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  arrange(padj) %>%
  mutate(
    significance = case_when(
      padj < 0.05 & log2FoldChange >  1 ~ "Up",
      padj < 0.05 & log2FoldChange < -1 ~ "Down",
      TRUE ~ "NS"  # Not Significant
    ),
    significance = factor(significance, levels = c("Up", "Down", "NS"))
  )


table(res_df$significance)

```



```{r}

res_df %>% filter(significance == "Up") %>% head(10) %>%
  dplyr::select(gene, log2FoldChange, padj) %>% print()

```

VST

```{r}
counts_gene_1 <- counts_gene[,-1]

dds_all <- DESeqDataSetFromMatrix(
  countData = counts_gene_1[rowSums(counts_gene_1) >= 10, ],
  colData   = meta %>% column_to_rownames("Sample"),
  design    = ~ Type)


vst_all <- vst(dds_all, blind = TRUE)

```

PCA Plot

```{r}

vst_pca <- prcomp(vst_all$Counts,
                   scale=TRUE)
vst_pca$Type <-  vst_all$Type
PC <- as.data.frame(vst_pca$x)

PC$Type <- vst_all$Type

row.names(PC) <- c("SRR11647685", "SRR11647687", "SRR11647689", "SRR11647695", "SRR11647697", "SRR11647699")
head(PC)
summary(PC)



```

```{r}

ggplot(PC, aes(x=Type, y=PC1)) +
  geom_point()+
  geom_text(
    label=rownames(PC), 
    nudge_x = 0.2, nudge_y = 0.05, 
    check_overlap = F
  )+
  ggtitle("PCA of VST- normalized counts")


```

Sample-to-Sample Correlation Heatmap

```{r}

vst_mat <- assay(vst_all)
cor_mat  <- cor(vst_mat, method = "pearson")


anno_df <- meta %>%
  column_to_rownames("Sample") %>%
  dplyr::select(Type)


anno_colors <- list(
  Type = c(
    Tumor                 = "#96C",
    Normal                 = "#16A"
  )
)


pheatmap(
  cor_mat,
  annotation_col = anno_df,
  annotation_row = anno_df,
  annotation_colors = anno_colors,
  color = colorRampPalette(c("yellow", "red"))(50),
  border_color = NA,
  main = "Sample-to-Sample Pearson Correlation",
  fontsize = 10
)
```

Heatmap of Top Differentially Expressed Genes

```{r}

top_genes <- res_df %>%
  filter(!is.na(padj)) %>%
  arrange(padj) %>%
  head(50) %>%
  pull(gene)

top_genes <-as.numeric(top_genes)

samples_li_unt <- meta %>%
  filter(Type %in% c("Tumor", "Normal")) %>%
  pull(Sample)


vst_subset <-  vst_mat[top_genes,]
vst_z <- t(scale(t(vst_subset)))

anno_subset <- meta %>%
  filter(Sample %in% samples_li_unt) %>%
  column_to_rownames("Sample") %>%
  dplyr::select(Type)

sum(is.na(vst_z))

pheatmap(
  vst_z,
  annotation_col  = anno_subset,
  annotation_colors = list(Typr = anno_colors$Type[c("Normal", "Tumor")]),
  color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(50),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = TRUE,
  fontsize_row = 8,
  border_color = NA,
  main = "Top 50 DE Genes — Tumor vs. Normal (Z-scored VST)"
)
```

MA Plot

```{r}

plotMA(dds,
  main = "MA plot - Tumor vs Normal Tissue")
```


Volcano Plot

```{r}

ggplot(data = res_df, aes(x = log2FoldChange, y = -log10(padj), col = significance)) +
  geom_vline(xintercept = c(-1, 1), col = "gray", linetype = 'dashed') +
  geom_hline(yintercept = -log10(0.5), col = "gray", linetype = 'dashed') +
  geom_point(size = 1)+
  scale_color_manual(values = c("#00AFBB","#FFDB6D", "grey" ), 
                     labels = c("Upregulated", "Downregulated", "Not significant"))+
   labs(title = "Volcano Plot - Tumor vs Normal",
     subtitle = "padj < 0.05 and |log2FC| > 1 threshhold shown as dashed lines",
     x ="log2 Fold Change (Normal/Tumor", 
     y = "-log10(adjusted p-value)")


```

Gene List Preparation

```{r}

sig_up   <- res_df %>% filter(significance == "Up")   %>% pull(gene)
sig_down <- res_df %>% filter(significance == "Down")  %>% pull(gene)
sig_all  <- res_df %>% filter(significance != "NS")    %>% pull(gene)

cat("Upregulated:", length(sig_up), "genes\n")
cat("Downregulated:", length(sig_down), "genes\n")
cat("Total significant:", length(sig_all), "genes\n")
```

```{r}

ranked_list <- res_df %>%
  filter(!is.na(log2FoldChange), !is.na(padj)) %>%
  arrange(desc(log2FoldChange)) %>%
  dplyr::select(gene, log2FoldChange)

counts_gene <- tibble::rownames_to_column(counts_gene, "gene")

count_name <- data.frame(gene = counts_gene$gene,
                   Geneid = counts_gene$Geneid)


ranked_list <- merge(ranked_list, count_name, by='gene', all.x=TRUE)



gene_ranks <- setNames(ranked_list$log2FoldChange, ranked_list$Geneid)
cat("\nRanked list length:", length(gene_ranks), "\n")
cat("Range of log2FC:", round(range(gene_ranks), 2), "\n")


```
Converting Gene Symbols to ENTREZ IDs

```{r}
symbol_to_entrez <- bitr(
  gsub("\\..*$", "", names(gene_ranks)),
  fromType = "ENSEMBL",
  toType   = c("ENTREZID", "SYMBOL"),
  OrgDb    = org.Hs.eg.db
)

cat("Mapped", nrow(symbol_to_entrez), "of", length(gene_ranks), "genes to Entrez IDs\n")
head(counts_gene)
```

```{r}


ranked_entrez <- gene_ranks[symbol_to_entrez$ENSEMBL]
names(ranked_entrez) <- symbol_to_entrez$ENTREZID

ranked_entrez <- sort(ranked_entrez, decreasing = TRUE)
ranked_entrez <- ranked_entrez[!duplicated(names(ranked_entrez))]


ranked_symbol <- gene_ranks[symbol_to_entrez$ENSEMBL]
names(ranked_symbol) <- symbol_to_entrez$SYMBOL

ranked_symbol <- sort(ranked_symbol, decreasing = TRUE)
ranked_symbol <- ranked_symbol[!duplicated(names(ranked_symbol))]

```

Over-Representation Analysis (ORA) — GO

```{r}
gene_list <- c(ranked_list$log2FoldChange, ranked_list$Geneid)

go_enrich <- enrichGO(gene = gene_list,
                      OrgDb = org.Hs.eg.db, 
                      keyType = 'ENSEMBL',
                      readable = T,
                      ont = "BP",
                      pvalueCutoff = 0.05, 
                      qvalueCutoff = 0.10)

```

```{r}

barplot(go_enrich, 
        drop = TRUE, 
        showCategory = 10, 
        title = "GO Biological Process - ORA (Tumor vs Normal)",
        font.size = 8)
```
```{r}
dotplot(go_enrich,
  title="GO Biological Process - ORA Droplot")

```

ORA — KEGG Pathways


```{r}
kegg_list <- c(ranked_list$log2FoldChange, ranked_list$gene)

kegg_organism = "hsa"
kk <- enrichKEGG(gene=kegg_list, universe=names(kegg_list),organism=kegg_organism, pvalueCutoff = 0.05, keyType = "ncbi-geneid")


dotplot(kk, 
        showCategory = 10, 
        title = "KEGG Pathways- ORA (Tumor vs Normal",
        font.size = 8)
```

ORA — MSigDB Hallmark Gene Sets

```{r}

hallmark_sets <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, ncbi_gene)


msig_ora <- enricher(
  gene     = res_df$gene,
  TERM2GENE = hallmark_sets,
  pvalueCutoff = 0.05
)

if (nrow(msig_ora@result) > 0) {
  dotplot(msig_ora,
          showCategory = 20,
          title = "MSigDB Hallmark ORA — Normal vs Tumor") +
    theme(axis.text.y = element_text(size = 10))
} else {
  cat("No significant Hallmark terms found at padj < 0.05.\n")

  msig_ora_all <- enricher(gene = sig_all, TERM2GENE = hallmark_sets, pvalueCutoff = 1)
  dotplot(msig_ora_all, showCategory = 15,
          title = "MSigDB Hallmark (all terms, unfiltered)") +
    theme(axis.text.y = element_text(size = 10))
}

```

Gene Set Enrichment Analysis (GSEA) — GO and GSEA Enrichment Plot

```{r}

gene_list_1 <- ranked_list$log2FoldChange
names(gene_list_1) <- ranked_list$Geneid
gene_list_1=sort(gene_list_1, decreasing = TRUE)


gse <- gseGO(geneList=gene_list_1, 
             ont ="ALL", 
             keyType = "ENSEMBL", 
             nPerm = 10000, 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = org.Hs.eg.db, 
             pAdjustMethod = "none")

gseaplot2(gse, title = "GSEA Enrichment - Top GO Terms", geneSetID = 1)
```

GSEA Ridgeplot

```{r}
gse
ridgeplot(gse)+
  labs(title = "GSEA GO - Distribution of Log2FC for Top Pathways")

```

GSEA — KEGG

```{r}
gene_list_2 <- ranked_list$log2FoldChange
names(gene_list_2) <- ranked_list$gene

gene_list_2=sort(gene_list_2, decreasing = TRUE)

GSEA_KEGG <- gseKEGG(geneList = gene_list_2, 
                     seed = TRUE)  


dotplot(GSEA_KEGG, 
        showCategory = 10, 
        title = "GSEA KEGG Pathways - Tumor vs Normal",
        font.size = 8)
```
